An Improved Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2021-01-07 simplified_sdc_intro/paper.tex
some updates based on the current code
some updates

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in with helium burning for
temperatures in excess of $4\times 10^9~\mathrm{K}$.  Physically, the
forward and reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium.  These large changes in abundances (and large alternately
positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep, but this makes simulations prohibitively expensive.
Sometimes the burning is simply halted on a zone-by-zone basis when
NSE conditions are reached (e.g., as in \citealt{hedet}).
Alternately, at high temperatures, a reaction network can be replaced
with a table of NSE abundances and the zone's composition set through
table look-ups (e.g.\ \citealt{ma:2013}) \MarginPar{need more here}

@@

The compositional quantities
it carries, $\bar{A}$ and $Y_e$ are not representable from the 19
isotopes we carry in the main network. For this reason, when we are
using the NSE network, we always provide these two composition quantities as
inputs to the EOS rather than using the $X_k$ directly.
the EOS directly from the auxiliary state in each zone instead of
using the $X_k$ directly.
Our equation of state needs $\bar{Z}$ in addition
to the auxiliary quantities, which is is easily computed as
\begin{equation}
\bar{Z} = \bar{A} Y_e
\end{equation}

@@

For Strang splitting, the hydrodynamic and reactive substeps over
the overall time-advancemenet scheme using the aprox19 + NSE network
are as follows:
\begin{itemize}

@@

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:nse_strang_methodII} Convergence ($L_1$ norm) for the NSE convergence
problem using Strang splitting.}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $4.522 \times 10^{19}$  & 1.659  & $1.432 \times 10^{19}$  & 1.542  & $4.917 \times 10^{18}$  \\
 $\rho u$                    & $1.883 \times 10^{28}$  & 1.681  & $5.873 \times 10^{27}$  & 1.524  & $2.042 \times 10^{27}$  \\
 $\rho v$                    & $1.883 \times 10^{28}$  & 1.681  & $5.873 \times 10^{27}$  & 1.524  & $2.042 \times 10^{27}$  \\
 $\rho E$                    & $5.933 \times 10^{37}$  & 1.613  & $1.940 \times 10^{37}$  & 1.411  & $7.293 \times 10^{36}$  \\
 $\rho e$                    & $5.931 \times 10^{37}$  & 1.613  & $1.939 \times 10^{37}$  & 1.411  & $7.291 \times 10^{36}$  \\
 $T$                         & $9.885 \times 10^{20}$  & 1.207  & $4.281 \times 10^{20}$  & 1.058  & $2.057 \times 10^{20}$  \\
 $\rho Y_e$                  & $2.379 \times 10^{19}$  & 1.678  & $7.431 \times 10^{18}$  & 1.578  & $2.489 \times 10^{18}$  \\
 $\rho \bar{A}$              & $5.141 \times 10^{21}$  & 1.230  & $2.191 \times 10^{21}$  & 1.087  & $1.032 \times 10^{21}$  \\
 $\rho (B/A)$                & $4.149 \times 10^{20}$  & 1.693  & $1.283 \times 10^{20}$  & 1.562  & $4.345 \times 10^{19}$  \\
\enddata
\end{deluxetable}

>>>>

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in for
temperatures in excess of $\mbox{few} \times 10^9~\mathrm{K}$.  Physically, the
forward and reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium.  These large changes in abundances (and large alternately
positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep, but this makes simulations prohibitively expensive.
Sometimes the burning is simply halted on a zone-by-zone basis when
NSE conditions are reached (e.g., as in \citealt{hedet}).
Alternately, at high temperatures, a reaction network can be replaced
with a table of NSE abundances and the zone's composition set through
table look-ups (e.g.\ \citealt{ma:2013}) \MarginPar{need more here}

@@

The compositional quantities
it carries, $\bar{A}$ and $Y_e$ are not representable from the 19
isotopes we carry in the main network. For this reason, when we are
using the NSE network, we always provide these two composition quantities as
inputs to the EOS rather than using the $X_k$ directly.
the EOS directly from the auxiliary state in each zone instead of
using the $X_k$ directly.
Our equation of state needs the mean charge per nucleus, $\bar{Z}$, in addition
to the auxiliary quantities, which is computed as
\begin{equation}
\bar{Z} = \bar{A} \sum_k \frac{Z_k X_k}{A_k} = \bar{A} Y_e
\end{equation}
(see, e.g., \citealt{flash}).

@@

For Strang splitting, we alternate the reaction and hydrodynamics,
with each operation working on the output of the previous operation.
If we define an advection operator over a timestep $\Delta t$ acting on $\Uc$ as
$\mathbb{A}_{\Delta t}(\Uc)$ and a reaction operator over a timestep of
$\Delta t/2$ acting on $\Uc$ as $\mathbb{R}_{\Delta t/2}(\Uc)$, then the advance appears
as:
\begin{align}
  \Uc^\star &= \mathbb{R}_{\Delta t/2}(\Uc^n) \\
  \Uc^{n+1,\star} &= \mathbb{A}_{\Delta t}(\Uc^\star) \\
  \Uc^{n+1} &= \mathbb{R}_{\Delta t/2}(\Uc^{n+1,\star})
\end{align}
Tthe hydrodynamic and reactive substeps over
the overall time-advancemenet scheme using the aprox19 + NSE network
are as follows:
\begin{itemize}

@@

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:nse_strang_methodII} Convergence ($L_1$ norm) for the NSE convergence
problem using Strang splitting.}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $4.522 \times 10^{19}$  & 1.659  & $1.432 \times 10^{19}$  & 1.542  & $4.917 \times 10^{18}$  \\
 $\rho u$                    & $1.883 \times 10^{28}$  & 1.681  & $5.873 \times 10^{27}$  & 1.524  & $2.042 \times 10^{27}$  \\
 $\rho v$                    & $1.883 \times 10^{28}$  & 1.681  & $5.873 \times 10^{27}$  & 1.524  & $2.042 \times 10^{27}$  \\
 $\rho E$                    & $5.933 \times 10^{37}$  & 1.613  & $1.940 \times 10^{37}$  & 1.411  & $7.293 \times 10^{36}$  \\
 $\rho e$                    & $5.931 \times 10^{37}$  & 1.613  & $1.939 \times 10^{37}$  & 1.411  & $7.291 \times 10^{36}$  \\
 $T$                         & $9.885 \times 10^{20}$  & 1.207  & $4.281 \times 10^{20}$  & 1.058  & $2.057 \times 10^{20}$  \\
 $\rho X(\isotm{He}{4})$     & $5.086 \times 10^{17}$  & 1.430  & $1.887 \times 10^{17}$  & 1.187  & $8.288 \times 10^{16}$  \\
 $\rho X(\isotm{C}{12})$     & $1.585 \times 10^{12}$  & 1.829  & $4.462 \times 10^{11}$  & 1.433  & $1.652 \times 10^{11}$  \\
 $\rho X(\isotm{O}{16})$     & $8.060 \times 10^{12}$  & 1.706  & $2.470 \times 10^{12}$  & 1.347  & $9.710 \times 10^{11}$  \\
 $\rho X(\isotm{Cr}{48})$    & $1.149 \times 10^{19}$  & 1.400  & $4.353 \times 10^{18}$  & 1.258  & $1.820 \times 10^{18}$  \\
 $\rho X(\isotm{Fe}{52})$    & $9.018 \times 10^{19}$  & 1.173  & $4.001 \times 10^{19}$  & 1.127  & $1.832 \times 10^{19}$  \\
 $\rho X(\isotm{Fe}{54})$    & $5.036 \times 10^{20}$  & 1.731  & $1.517 \times 10^{20}$  & 1.727  & $4.582 \times 10^{19}$  \\
 $\rho X(\isotm{Ni}{56})$    & $5.686 \times 10^{20}$  & 1.622  & $1.847 \times 10^{20}$  & 1.563  & $6.252 \times 10^{19}$  \\
 $\rho Y_e$                  & $2.379 \times 10^{19}$  & 1.678  & $7.431 \times 10^{18}$  & 1.578  & $2.489 \times 10^{18}$  \\
 $\rho \bar{A}$              & $5.141 \times 10^{21}$  & 1.230  & $2.191 \times 10^{21}$  & 1.087  & $1.032 \times 10^{21}$  \\
 $\rho (B/A)$                & $4.149 \times 10^{20}$  & 1.693  & $1.283 \times 10^{20}$  & 1.562  & $4.345 \times 10^{19}$  \\
\enddata
\end{deluxetable}

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:nse_strang_methodII} Convergence ($L_1$ norm) for the NSE convergence
problem using the simplified-SDC algorithm.}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $4.558 \times 10^{19}$  & 1.633  & $1.470 \times 10^{19}$  & 1.578  & $4.923 \times 10^{18}$  \\
 $\rho u$                    & $1.945 \times 10^{28}$  & 1.648  & $6.208 \times 10^{27}$  & 1.513  & $2.176 \times 10^{27}$  \\
 $\rho v$                    & $1.945 \times 10^{28}$  & 1.648  & $6.208 \times 10^{27}$  & 1.513  & $2.176 \times 10^{27}$  \\
 $\rho E$                    & $5.710 \times 10^{37}$  & 1.620  & $1.857 \times 10^{37}$  & 1.552  & $6.333 \times 10^{36}$  \\
 $\rho e$                    & $5.708 \times 10^{37}$  & 1.620  & $1.857 \times 10^{37}$  & 1.553  & $6.331 \times 10^{36}$  \\
 $T$                         & $5.873 \times 10^{20}$  & 1.621  & $1.910 \times 10^{20}$  & 1.494  & $6.779 \times 10^{19}$  \\
 $\rho X(\isotm{He}{4})$     & $2.453 \times 10^{17}$  & 1.835  & $6.876 \times 10^{16}$  & 1.657  & $2.181 \times 10^{16}$  \\
 $\rho X(\isotm{C}{12})$     & $1.130 \times 10^{12}$  & 1.846  & $3.145 \times 10^{11}$  & 1.678  & $9.831 \times 10^{10}$  \\
 $\rho X(\isotm{O}{16})$     & $5.339 \times 10^{12}$  & 1.830  & $1.502 \times 10^{12}$  & 1.667  & $4.730 \times 10^{11}$  \\
 $\rho X(\isotm{Cr}{48})$    & $6.880 \times 10^{18}$  & 1.791  & $1.988 \times 10^{18}$  & 1.601  & $6.554 \times 10^{17}$  \\
 $\rho X(\isotm{Fe}{52})$    & $3.470 \times 10^{19}$  & 1.606  & $1.140 \times 10^{19}$  & 1.608  & $3.740 \times 10^{18}$  \\
 $\rho X(\isotm{Fe}{54})$    & $4.766 \times 10^{20}$  & 1.764  & $1.403 \times 10^{20}$  & 1.777  & $4.093 \times 10^{19}$  \\
 $\rho X(\isotm{Ni}{56})$    & $4.835 \times 10^{20}$  & 1.781  & $1.407 \times 10^{20}$  & 1.788  & $4.075 \times 10^{19}$  \\
 $\rho Y_e$                  & $2.370 \times 10^{19}$  & 1.644  & $7.582 \times 10^{18}$  & 1.580  & $2.536 \times 10^{18}$  \\
 $\rho \bar{A}$              & $4.061 \times 10^{21}$  & 1.678  & $1.270 \times 10^{21}$  & 1.675  & $3.975 \times 10^{20}$  \\
 $\rho (B/A)$                & $4.224 \times 10^{20}$  & 1.664  & $1.333 \times 10^{20}$  & 1.602  & $4.391 \times 10^{19}$  \\
\enddata
\end{deluxetable}
